Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-06 paper.tex
Fixup refs

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{makecell}

@@

Recombination via template switching is a common feature in the evolution of
coronaviruses (CoV) \cite{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
\cite{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with CoV adaptation to new host species and with the production of
more immune evasive variants \cite{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
to gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC) harbouring multiple characteristic
mutations increased the detectability of recombination. By searching for
genomes containing segments from different VoC, multiple recombinant lineages
have been detected, for example, between Alpha and pre-VoC
\cite{Jackson2021-ik}, Alpha and Epsilon \cite{Wertheim2022-hj}, and Alpha and
Delta \cite{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\cite{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
transmission (more than five descendants) and monophyly (> 70\% bootstrap
support) are eligible for a Pango lineage designation \cite{Rambaut2020-dw},
starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
Naming Recombinant Lineages” 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages (e.g.,
\cite{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\cite{Ignatieva2021-rg}. More recently, Turakhia et al. \cite{Turakhia2022-it}
has presented a phylogeny-based approach ("RIPPLES") to identify putative
recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying
motifs. To find candidate recombinants, RIPPLES first builds a phylogeny
without assuming recombination and then scans for long branches (i.e., carrying
many mutations) and then determines if these candidates would be better
explained by recombination. Specifically, RIPPLES exhaustively breaks each
candidate sequence into segments and attempts to find more parsimonious
placements for each segment on the phylogeny. If such placements are found, the
sequence is identified as a putative recombinant. Although it enables rapid
searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
phylogeny that accounts for only mutations, treats recombinants \textit{post
hoc}, and is an incomplete representation of the reticulate evolutionary
history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
of recombination is possible when recombinant lineages are rare and leave few
descendants, but this is becoming increasingly untenable with the proliferation
of recombinant lineages (e.g., more than half of the sequences sampled in
February, 2023 are from XBB and its descendants; \cite{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing challenge. This challenge is particularly
acute for SARS-CoV-2, given the large amount of data that has been shared, with
over 15 million sequences in the GISAID database \cite{Shu2017-hp}. It is well
known that recombination distorts phylogenies \cite{Schierup2000-fg} and
affects the results of downstream analysis (e.g., inference of selection;
\cite{Anisimova2003-vr}). However, standard phylogenetic methods do not account
for recombination (e.g., \cite{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
Alternatively, genetic ancestries of recombining lineages can be represented by
ancestral recombination graphs (ARG; \cite{Griffiths1981-lw,Hudson1983-if},
which  capture network-like phylogenetic relationships \cite{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human
genomes and beyond
\cite{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
Wohns et al. \cite{Wohns2022-th} reconstructed a unified whole-genome genealogy
of thousands of modern and ancient human genomes, based on over 90 million
variable sites. To achieve this, Wohns et al. \cite{Wohns2022-th} utilised
several computational advances in ARG inference: (1) the "succinct tree
sequence" data structure that can compactly store large ARGs
\cite{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
infer copying paths between genomes \cite{Kelleher2019-ba}; and (3) efficient,
open-source software tools
\cite{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(“sc2ts”, pronounced “scoots”) that builds on the advances above and
additionally takes advantage of known collection dates to infer ARGs from
millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
information about SARS-CoV-2 that can be efficiently accessed using "tskit"
(tree sequence toolkit) software library \cite{Kelleher2018-xc}. Sc2ts is a
promising method to automate detection of SARS-CoV-2 recombinants and to
jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at large scale.

@@

The findings of this study are based on sequences and metadata available on
GISAID (\href{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0 \cite{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a “bad” quality control status
in any of the four Nextclade columns ("qc.missingData.status”,
“qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
These excluded genomes are enriched for low-coverage, contaminated, or
misassembled sequences.

@@

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
genome alignments were encoded as N. Problematic bases in the alignments, which
had two or more ambiguous letters or gaps within a distance of seven bases,
were masked as missing data following the protocol in
\href{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
when attaching a sample genome to the ARG, but they were subsequently imputed.
Further, 481 problematic sites flagged as prone to sequencing errors or highly
homoplasic were excluded entirely
(\href{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\href{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

@@

We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
efficient HMM implementation of the Li and Stephens (LS) model \cite{Li2003-ib}
to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
collection dates of the SARS-CoV-2 samples to determine the order of genome
copying; and (3) several parsimony-based heuristics to locally refine the
topology. In brief, this combined approach uses the likelihood model of LS to
attach sample genomes to existing nodes and to detect recombinants, but because
many of the tip samples attached daily harbour few, if any, mutational changes,
we use parsimony to resolve the relationships near the tip samples.

@@

Each daily batch of genomes is first attached to the growing ARG without any
phylogenetic resolution. In cases where the newly added sequence is inferred to
be a recombinant, a recombination node is inserted; the “date” of this node is
arbitrarily set to the average of the collection dates of the genome and the
most recent of its parent nodes. Otherwise, the genomes are attached to the
most closely related ancestral node (an attachment node). This creates
polytomies that are then locally resolved among the descendants of each
attachment node using the Unweighted Pair Group Method with Arithmetic Mean
(UPGMA) algorithm \cite{Michener1957-tr}, with average linkage (Figure 1C).
This hierarchical clustering algorithm produces a dendrogram relating a set of
sequences, which is treated as an ultrametric binary tree here. Subsequently,
mutations are placed onto the daily tree by maximum parsimony. An issue with
the UPGMA algorithm is that it may create internal nodes supported by no
informative site (i.e., having no mutation immediately ancestral to them). In
an adjustment step, such unsupported internal nodes are removed to “un-resolve”
the daily tree; the remaining internal nodes, which are supported by at least
one shared mutation, are retained. The “dates” of the non-sample internal nodes
are set to the same date as the samples in the daily batch. The resulting daily
tree is grafted onto the attachment node in the global ARG.

@@

Furthermore, during inference, the ARG is automatically encoded, stored, and
updated as a tree sequence using tskit \cite{Kelleher2018-xc}.

@@

The LS model is a convenient probabilistic approximation to the coalescent with
recombination \cite{Hudson1983-if}, itself a generative model of a genealogical
process with recombination \cite{Li2003-ib}. Importantly, it can be implemented
as an HMM in which a focal genome is a sequence of nucleotides (observed
states) that are probabilistically emitted as an imperfect mosaic of a set of
reference genomes (hidden states). As such, an efficient HMM machinery can be
employed to make statements about parameters of interest and highlight
potential sources of error in genotyping and sequencing data. The LS model is
used in a wide variety of applications in genomics. For example, it is the
engine that drives statistical genotype phasing and imputation
\cite{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
\cite{Hinch2011-tz}), and selection within and across hosts in viral sequence
data (e.g., \cite{Palmer2019-wa}).

The generative process of the LS model is summarised in Figure 2. Here, a
transition matrix, Q, governs the process of switching (recombining) between
members of the reference panel (the hidden states). An emission matrix,
\textit{E}, allows for differences between the focal sequence and the hidden
state from which it is copied (due to mutation as well as genotyping or
sequencing error). Then, we use the Viterbi algorithm \cite{Viterbi1967-ol} to
find the most likely copying path through the sequences, given a set of
reference sequences, \textit{Q}, and \textit{E}.

@@

Non-sample nodes inserted during ARG inference had no Pango lineage assignment
from Nextclade. We imputed the Pango lineage of non-sample nodes using a
two-step procedure (Figure 3) to help assess how consistently breakpoint
intervals are estimated by the HMM runs (see below). If the Pango lineage of
the parental node was available and no mutation was immediately ancestral to
the non-sample node, then the parental node’s status was directly propagated to
the non-sample node. Otherwise, the Pango lineage of the non-sample node was
inferred by matching its full set of mutations against the Pango
lineage-defining mutations (based on 90\% consensus of the sequences analysed)
from the COVID-CG website \cite{Chen2021-zc}; \href{https://covidcg.org/};
accessed on November 04, 2022).

@@

The source code implementing the sc2ts pipeline is available via GitHub
\href{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
various functions and algorithms implemented in tskit and tsinfer
\href{https://github.com/tskit-dev/}. The pipeline uses the SciPy
implementation of the UPGMA algorithm to build daily trees
\href{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
The Jupyter notebooks used to produce the results described herein are also
available on GitHub \href{https://github.com/jeromekelleher/sc2ts-paper/}.

@@

For validation purposes, we compare the genetic relationships in the
sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
project \cite{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
restricted to non-recombinant lineages. To allow for comparison, the ARGs were
"simplified" \cite{Kelleher2018-xc} to retain only those tips of our genealogy
that match the samples present in a specific NextStrain tree. These simplified
"backbone" ARGs exhibit very little recombination (6 and 5 trees across the
genome for the wide and long ARGs, respectively).

@@

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by Turakhia et al.
\cite{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
the number of intersecting intervals in the N gene following the peak in ORF8
towards the 3’ end.

Next, we investigate the potential recombination signals alongside the number
of mutations across the genome (Figure 4C). Regions with a higher density of
mutations contain more genetic variation, increasing the power to detect
recombination events. The position 28,271 (located in the ORF8-N intergenic
region) has a particularly high number of mutations in the samples (Figure 5).
At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
(n = 1,605) appear to occur frequently. One possible explanation is that this
homopolymeric region is prone to sequencing errors, giving rise to an
artifactual recombination signal. At the upstream position 27,972, numerous
mutations also occur. This position has been reported to be hypermutable
\cite{Jungreis2021-dh}.

We further consider the breakpoint intervals within the context of
recombination breakpoint sequence motifs. It is hypothesised that certain
palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
template switching during replication in SARS-CoV-2 via formation of
base-paired stem loops in the genome structure \cite{Gallaher2020-lb}. There
are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
breakpoint intervals of the HMM-consistent recombinants span at least one of
the breakpoint sequences.

@@

In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. Yi et al. \cite{Yi2021-sc} reconstructed a SARS-CoV-2
phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
January 12, 2021, and systematically catalogued all the mutations occurring
along the phylogeny. Notably, in the mutational spectrum described by Yi et
al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
mutations, respectively.

For validation purposes, we compare the mutational spectrum from the wide ARG
with the mutational spectrum of Yi et al. We categorised all of the single
nucleotide mutations by type (defined by the ancestral state and derived
state), excluded mutations inherited by only a single sample (which are
enriched for sequencing errors and artefacts), and tallied them up by type.
Similarly, we took the data for single nucleotide mutations from Yi et al.
\ref{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
occurring along terminal branches, and tallied them up by type. We find that
the mutational spectrum from the wide ARG (448,825 mutations) matches the
mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
including the mutations inherited by only a single sample or occurring on
terminal branches (data not shown).

@@

\begin{figure} \centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
\cite{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C > U) and the
lighter colours the reverse (e.g., U > C).} \end{figure}

\bibliographystyle{alpha}
\bibliography{paper}

>>>>

\usepackage{authblk}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{url}
% JK: turning this off for the moment as I keep clicking through on links
% to the bibliography while reading the text and it's intensely annoying.
% Can reinstate when we're ready to preprint
% \usepackage[hidelinks]{hyperref}

\usepackage{makecell}

@@

Recombination via template switching is a common feature in the evolution of
coronaviruses (CoV) \citep{Graham2010-xe,De_Klerk2022-tt}), including SARS-CoV-2
\citep{VanInsberghe2021-eu,Jackson2021-ik,Ignatieva2021-rg}. By bringing
together mutations carried by different lineages, recombination plays an
important role in generating genomic diversity, with recombinant lineages
associated with CoV adaptation to new host species and with the production of
more immune evasive variants \citep{Graham2010-xe,De_Klerk2022-tt}.

Recombination was challenging to detect in SARS-CoV-2 early in the pandemic due
to gradual accumulation of genetic variation. In late 2020, the appearance and
spread of variants of concern (VoC) harbouring multiple characteristic
mutations increased the detectability of recombination. By searching for
genomes containing segments from different VoC, multiple recombinant lineages
have been detected, for example, between Alpha and pre-VoC
\citep{Jackson2021-ik}, Alpha and Epsilon \citep{Wertheim2022-hj}, and Alpha and
Delta \citep{Sekizuka2022-xz}). More recently, the high prevalence of Omicron,
with multiple co-circulating deeply divergent lineages (BA.1 to BA.5), has
accelerated the rate of coinfection and the potential for recombination
\citep{Bal2022-hq}. Recombinant lineages that have sufficient evidence of onward
transmission (more than five descendants) and monophyly (> 70\% bootstrap
support) are eligible for a Pango lineage designation \citep{Rambaut2020-dw},
starting with the letter X (“Pango Lineage Nomenclature: Provisional Rules for
Naming Recombinant Lineages” 2021).

Most genomic surveys for SARS-CoV-2 recombinants search for mosaic genomes that
combine specific subsets of mutations from different lineages (e.g.,
\citep{VanInsberghe2021-eu,Jackson2021-ik,Wertheim2022-hj,Sekizuka2022-xz}).
This approach only identifies those inter-lineage recombination events
resulting in the patterns investigated. Others have inferred recombinant
genealogies to disentangle recombination from recurrent mutation, without
having to pre-specify characteristic mutations for each lineage
\citep{Ignatieva2021-rg}. More recently, Turakhia et al. \citep{Turakhia2022-it}
has presented a phylogeny-based approach ("RIPPLES") to identify putative
recombinants among over ten million SARS-CoV-2 genomes, without pre-specifying
motifs. To find candidate recombinants, RIPPLES first builds a phylogeny
without assuming recombination and then scans for long branches (i.e., carrying
many mutations) and then determines if these candidates would be better
explained by recombination. Specifically, RIPPLES exhaustively breaks each
candidate sequence into segments and attempts to find more parsimonious
placements for each segment on the phylogeny. If such placements are found, the
sequence is identified as a putative recombinant. Although it enables rapid
searches for genomic evidence of recombinants, RIPPLES relies on a SARS-CoV-2
phylogeny that accounts for only mutations, treats recombinants \textit{post
hoc}, and is an incomplete representation of the reticulate evolutionary
history of SARS-CoV-2. As noted by the authors, a \textit{post hoc} treatment
of recombination is possible when recombinant lineages are rare and leave few
descendants, but this is becoming increasingly untenable with the proliferation
of recombinant lineages (e.g., more than half of the sequences sampled in
February, 2023 are from XBB and its descendants; \citep{Chen2022-pz}.

Inferring evolutionary trees while accounting for both mutation and
recombination is a long-standing challenge. This challenge is particularly
acute for SARS-CoV-2, given the large amount of data that has been shared, with
over 15 million sequences in the GISAID database \citep{Shu2017-hp}. It is well
known that recombination distorts phylogenies \citep{Schierup2000-fg} and
affects the results of downstream analysis (e.g., inference of selection;
\citep{Anisimova2003-vr}). However, standard phylogenetic methods do not account
for recombination (e.g., \citep{Ronquist2012-zw,Minh2020-lr,Guindon2003-zd}).
Alternatively, genetic ancestries of recombining lineages can be represented by
ancestral recombination graphs (ARG; \citep{Griffiths1981-lw,Hudson1983-if},
which  capture network-like phylogenetic relationships \citep{Gusfield2014-qw}.

Approaches to infer ARGs now exist that can scale to tens of thousands of human
genomes and beyond
\citep{Speidel2019-yh,Kelleher2019-ba,Zhang2021-lf,Schaefer2021-yg}. Recently,
Wohns et al. \citep{Wohns2022-th} reconstructed a unified whole-genome genealogy
of thousands of modern and ancient human genomes, based on over 90 million
variable sites. To achieve this, Wohns et al. \citep{Wohns2022-th} utilised
several computational advances in ARG inference: (1) the "succinct tree
sequence" data structure that can compactly store large ARGs
\citep{Kelleher2018-xc}; (2) a highly efficient Hidden Markov Model (HMM)
implementation of the Li and Stephens (LS) model (Li and Stephens 2003) to
infer copying paths between genomes \citep{Kelleher2019-ba}; and (3) efficient,
open-source software tools
\citep{Baumdicker2022-ep,Kelleher2018-xc,Kelleher2019-ba,Wohns2022-th}.

Here, we present a preliminary version of a novel scalable methodology
(“sc2ts”, pronounced “scoots”) that builds on the advances above and
additionally takes advantage of known collection dates to infer ARGs from
millions of full-length SARS-CoV-2 genomes. Using publicly available SARS-CoV-2
genome sequences, we demonstrate that sc2ts can be used to simultaneously (1)
detect putative SARS-CoV-2 recombinants; (2) infer the phylogenetic
relationships of the major SARS-CoV-2 lineages; (3) reconstruct the mutational
spectrum of SARS-CoV-2; and (4) provide a compact record of evolutionary
information about SARS-CoV-2 that can be efficiently accessed using "tskit"
(tree sequence toolkit) software library \citep{Kelleher2018-xc}. Sc2ts is a
promising method to automate detection of SARS-CoV-2 recombinants and to
jointly model recombinational and mutational processes shaping the genomic
diversity of SARS-CoV-2 at large scale.

@@

The findings of this study are based on sequences and metadata available on
GISAID (\url{https://gisaid.org/}) up to Aug 22, 2022 and accessible at
doi.org/10.55876/gis8.230329cd (Supplementary Table 1). Sequences were removed
if they (1) had ambiguous collection dates; (2) were collected before January
01, 2020 or after June 30, 2022; or (3) were isolated from a non-human host.
Sequences were aligned to the Wuhan-Hu-1/2019 reference sequence (GenBank:
MN908947.3) using Nextclade v2.3.0 \citep{Aksamentov2021-hj} (dataset tag
2022-07-26T12:00:00Z). To reduce the impact of poor-quality sequences on ARG
inference, sequences were excluded if they had a “bad” quality control status
in any of the four Nextclade columns ("qc.missingData.status”,
“qc.mixedSites.status”, “qc.frameShifts.status”, and “qc.stopCodons.status”).
These excluded genomes are enriched for low-coverage, contaminated, or
misassembled sequences.

@@

Ambiguous nucleotide letters (i.e., not A, C, G, T, or a gap) in the pairwise
genome alignments were encoded as N. Problematic bases in the alignments, which
had two or more ambiguous letters or gaps within a distance of seven bases,
were masked as missing data following the protocol in
\url{https://github.com/roblanf/sarscov2phylo/}. Masked bases were ignored
when attaching a sample genome to the ARG, but they were subsequently imputed.
Further, 481 problematic sites flagged as prone to sequencing errors or highly
homoplasic were excluded entirely
(\url{https://virological.org/t/issues-with-sars-cov-2-sequencing-data/473};
\url{https://raw.githubusercontent.com/W-L/ProblematicSites_SARS-CoV2/master/problematic_sites_sarsCov2.vcf}).

@@

We introduce an ARG inference method (schematic in Figure 1) that uses (1) an
efficient HMM implementation of the Li and Stephens (LS) model \citep{Li2003-ib}
to copy SARS-CoV-2 genomes (see Sections 2.2 and 2.3); (2) the reported
collection dates of the SARS-CoV-2 samples to determine the order of genome
copying; and (3) several parsimony-based heuristics to locally refine the
topology. In brief, this combined approach uses the likelihood model of LS to
attach sample genomes to existing nodes and to detect recombinants, but because
many of the tip samples attached daily harbour few, if any, mutational changes,
we use parsimony to resolve the relationships near the tip samples.

@@

Each daily batch of genomes is first attached to the growing ARG without any
phylogenetic resolution. In cases where the newly added sequence is inferred to
be a recombinant, a recombination node is inserted; the “date” of this node is
arbitrarily set to the average of the collection dates of the genome and the
most recent of its parent nodes. Otherwise, the genomes are attached to the
most closely related ancestral node (an attachment node). This creates
polytomies that are then locally resolved among the descendants of each
attachment node using the Unweighted Pair Group Method with Arithmetic Mean
(UPGMA) algorithm \citep{Michener1957-tr}, with average linkage (Figure 1C).
This hierarchical clustering algorithm produces a dendrogram relating a set of
sequences, which is treated as an ultrametric binary tree here. Subsequently,
mutations are placed onto the daily tree by maximum parsimony. An issue with
the UPGMA algorithm is that it may create internal nodes supported by no
informative site (i.e., having no mutation immediately ancestral to them). In
an adjustment step, such unsupported internal nodes are removed to “un-resolve”
the daily tree; the remaining internal nodes, which are supported by at least
one shared mutation, are retained. The “dates” of the non-sample internal nodes
are set to the same date as the samples in the daily batch. The resulting daily
tree is grafted onto the attachment node in the global ARG.

@@

Furthermore, during inference, the ARG is automatically encoded, stored, and
updated as a tree sequence using tskit \citep{Kelleher2018-xc}.

@@

The LS model is a convenient probabilistic approximation to the coalescent with
recombination \citep{Hudson1983-if}, itself a generative model of a genealogical
process with recombination \citep{Li2003-ib}. Importantly, it can be implemented
as an HMM in which a focal genome is a sequence of nucleotides (observed
states) that are probabilistically emitted as an imperfect mosaic of a set of
reference genomes (hidden states). As such, an efficient HMM machinery can be
employed to make statements about parameters of interest and highlight
potential sources of error in genotyping and sequencing data. The LS model is
used in a wide variety of applications in genomics. For example, it is the
engine that drives statistical genotype phasing and imputation
\citep{Delaneau2019-wl,Browning2021-cg,Browning2018-nk,Rubinacci2020-pa}, and it
can be used to estimate mutation rates (e.g., REF), recombination rates (e.g.,
\citep{Hinch2011-tz}), and selection within and across hosts in viral sequence
data (e.g., \citep{Palmer2019-wa}).

The generative process of the LS model is summarised in Figure 2. Here, a
transition matrix, Q, governs the process of switching (recombining) between
members of the reference panel (the hidden states). An emission matrix,
\textit{E}, allows for differences between the focal sequence and the hidden
state from which it is copied (due to mutation as well as genotyping or
sequencing error). Then, we use the Viterbi algorithm \citep{Viterbi1967-ol} to
find the most likely copying path through the sequences, given a set of
reference sequences, \textit{Q}, and \textit{E}.

@@

Non-sample nodes inserted during ARG inference had no Pango lineage assignment
from Nextclade. We imputed the Pango lineage of non-sample nodes using a
two-step procedure (Figure 3) to help assess how consistently breakpoint
intervals are estimated by the HMM runs (see below). If the Pango lineage of
the parental node was available and no mutation was immediately ancestral to
the non-sample node, then the parental node’s status was directly propagated to
the non-sample node. Otherwise, the Pango lineage of the non-sample node was
inferred by matching its full set of mutations against the Pango
lineage-defining mutations (based on 90\% consensus of the sequences analysed)
from the COVID-CG website \citep{Chen2021-zc}; \url{https://covidcg.org/};
accessed on November 04, 2022).

@@

The source code implementing the sc2ts pipeline is available via GitHub
\url{https://github.com/jeromekelleher/sc2ts/}. The sc2ts pipeline uses
various functions and algorithms implemented in tskit and tsinfer
\url{https://github.com/tskit-dev/}. The pipeline uses the SciPy
implementation of the UPGMA algorithm to build daily trees
\url{https://docs.scipy.org/doc/scipy/reference/generated/scipy.cluster.hierarchy.average.html}.
The Jupyter notebooks used to produce the results described herein are also
available on GitHub \url{https://github.com/jeromekelleher/sc2ts-paper/}.

@@

For validation purposes, we compare the genetic relationships in the
sc2ts-inferred ARGs with phylogenetic relationships inferred by the NextStrain
project \citep{Hadfield2018-ef}. The NextStrain trees consist of subsamples of
SARS-CoV-2 genomes, which over the time period of our ARGs are almost entirely
restricted to non-recombinant lineages. To allow for comparison, the ARGs were
"simplified" \citep{Kelleher2018-xc} to retain only those tips of our genealogy
that match the samples present in a specific NextStrain tree. These simplified
"backbone" ARGs exhibit very little recombination (6 and 5 trees across the
genome for the wide and long ARGs, respectively).

@@

There is a preponderance of breakpoint intervals spanning the 3’ end of the
genome (Figure 4A), as noted previously by Turakhia et al.
\citep{Turakhia2022-it}. Inspecting more closely, we find an enrichment in the
number of intersecting intervals in the E gene, ORF8, ORF8-N intergenic region,
and 3’ untranslated region (Figure 4B). Markedly, there is an abrupt drop in
the number of intersecting intervals in the N gene following the peak in ORF8
towards the 3’ end.

Next, we investigate the potential recombination signals alongside the number
of mutations across the genome (Figure 4C). Regions with a higher density of
mutations contain more genetic variation, increasing the power to detect
recombination events. The position 28,271 (located in the ORF8-N intergenic
region) has a particularly high number of mutations in the samples (Figure 5).
At this position, mutations (n = 7,572), insertions (n = 5,782), and deletions
(n = 1,605) appear to occur frequently. One possible explanation is that this
homopolymeric region is prone to sequencing errors, giving rise to an
artifactual recombination signal. At the upstream position 27,972, numerous
mutations also occur. This position has been reported to be hypermutable
\citep{Jungreis2021-dh}.

We further consider the breakpoint intervals within the context of
recombination breakpoint sequence motifs. It is hypothesised that certain
palindromic breakpoint sequences (specifically, CAGAC and CAGAT) promote
template switching during replication in SARS-CoV-2 via formation of
base-paired stem loops in the genome structure \citep{Gallaher2020-lb}. There
are 87 occurrences of these two breakpoint sequences in the Wuhan-Hu-1/2019
reference sequence (Supplementary Table 5). We observe that XX (XX\%) of the
breakpoint intervals of the HMM-consistent recombinants span at least one of
the breakpoint sequences.

@@

In addition to recombination events, the ARGs record mutation events along the
SARS-CoV-2 phylogeny. Yi et al. \citep{Yi2021-sc} reconstructed a SARS-CoV-2
phylogeny of over 350,000 genomes sampled globally from December 24, 2019, to
January 12, 2021, and systematically catalogued all the mutations occurring
along the phylogeny. Notably, in the mutational spectrum described by Yi et
al., C-to-U and G-to-U mutations occur more frequently than U-to-C and U-to-G
mutations, respectively.

For validation purposes, we compare the mutational spectrum from the wide ARG
with the mutational spectrum of Yi et al. We categorised all of the single
nucleotide mutations by type (defined by the ancestral state and derived
state), excluded mutations inherited by only a single sample (which are
enriched for sequencing errors and artefacts), and tallied them up by type.
Similarly, we took the data for single nucleotide mutations from Yi et al.
\url{https://github.com/ju-lab/SC2_evol_signature}, excluded mutations
occurring along terminal branches, and tallied them up by type. We find that
the mutational spectrum from the wide ARG (448,825 mutations) matches the
mutational spectrum reported by Yi et al. (92,344 mutations) (Figure X). In
both spectra, C-to-U mutations and G-to-U mutations occur more frequently than
U-to-C and U-to-G, respectively (Figure X). Similar results are obtained when
including the mutations inherited by only a single sample or occurring on
terminal branches (data not shown).

@@

\begin{figure} \centering
\includegraphics[width=0.8\textwidth]{figures/mutational_spectra.pdf}
\caption{\label{fig:mutational_spectra} Mutational spectrum of SARS-CoV-2 in
the wide ARG (sampled to mid-2021) compared to that reported by Yi et al.
\citep{Yi2021-sc}. Mutations are categorised by type (i.e., ancestral state >
derived state). The percentages of each mutation type from the wide ARG are
represented by blue bars and the percentages from Yi et al. by orange bars,
with the darker colours representing one direction (e.g., C > U) and the
lighter colours the reverse (e.g., U > C).} \end{figure}

\bibliographystyle{plainnat}
\bibliography{paper}
